
# Reproducible Research: Peer Assessment 1
Author
Rudolf Stolz  
Version 1.0.0  
Date 140609

## Loading and preprocessing the data
---------------------------------------------------------------------------------------------
### Load Library
```{r echo = TRUE}
library(ggplot2)
library(lubridate)
library(plyr)
```

---------------------------------------------------------------------------------------------
### Load Raw Data
**Note:**  
This File assumes, that the data file activity.zip is stored in the R Working Directory!!!  
```{r echo = TRUE}
activity <- read.csv(unz("activity.zip", "activity.csv"), header=TRUE, sep=",")
```

## What is mean total number of steps taken per day?
---------------------------------------------------------------------------------------------
### Make a histogram of the total number of steps taken each day  
```{r echo = TRUE}
data.steps.1 <- data.frame(steps=as.numeric(length(unique(activity$date))))
count <- 1
for (i in unique(activity$date))
    {        
        data.steps.1[count,"steps"] <- sum(activity[activity$date==i & !is.na(activity$steps), "steps"])
        count <- count + 1
    }

hist(data.steps.1$steps, col="red", main="Total Number of Steps taken each day", 
     xlab="Number of Steps taken each day")
```

---------------------------------------------------------------------------------------------
### Calculate and report the mean and median total number of steps taken per day
```{r echo = TRUE}
summary(data.steps.1)

```

## What is the average daily activity pattern?
---------------------------------------------------------------------------------------------
### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo = TRUE}
data.steps.2 <- data.frame(interval=as.numeric(length(unique(activity$interval))),
                           steps=as.numeric(length(unique(activity$interval))))

count <- 1
for (i in unique(activity$interval))
    {        
        data.steps.2[count,"interval"] <- i 
        data.steps.2[count,"steps"] <- mean(activity[activity$interval==i & !is.na(activity$steps), "steps"])
        count <- count + 1
    }
plot(x=data.steps.2$interval,
     y=data.steps.2$steps, 
     type="l", lty="solid", lwd=1,
     col="black", main="Average Number of Steps per 5-minutes time interval", 
     xlab="Time Interval [min]",
     ylab="Average Steps [steps]")
```

---------------------------------------------------------------------------------------------
### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo = TRUE}
max.steps.2    <- max(data.steps.2[,"steps"])
max.interval.2 <- data.steps.2[which.max(data.steps.2[,"steps"]), "interval"]
```
The 5-minute interval **`r max.interval.2`** contains the maximum average number of 
steps of **`r max.steps.2`**

## Imputing missing values
---------------------------------------------------------------------------------------------
### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r echo = TRUE}
missing.values <- sum(is.na(activity$steps))
```
The number of missing step values is **`r missing.values`**

---------------------------------------------------------------------------------------------
### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

**I used the 5-minutes interval average to fill in the missing step values**   
**The required data are available from the calcuation of the last step and stored in data.steps.2**   
```{r echo = TRUE}
## calculate the new data set
activity.new <- activity
for (i in seq(1, nrow(activity.new)))
    {
        if (is.na(activity.new[i,"steps"]))
                {
                    newvalue <- data.steps.2[data.steps.2$interval==activity.new[i,"interval"], "steps"]
                    activity.new[i, "steps"] <- newvalue  
                }
    }
## calculate the histogram
data.steps.3 <- data.frame(steps=as.numeric(length(unique(activity.new$date))))
count <- 1
for (i in unique(activity.new$date))
    {        
        data.steps.3[count,"steps"] <- sum(activity.new[activity.new$date==i & !is.na(activity.new$steps), "steps"])
        count <- count + 1
    }

hist(data.steps.3$steps, col="red", main="Total Number of Steps taken each day",
     xlab="Number of Steps taken each day")

```
---------------------------------------------------------------------------------------------
### Calculate and report the mean and median total number of steps taken per day
```{r echo = TRUE}
summary(data.steps.3)

```
---------------------------------------------------------------------------------------------
### Compare the first and the last measurement
```{r echo = TRUE}
compare <- data.frame(summary(data.steps.1),summary(data.steps.3))
names(compare) <- c("col1", "col2", "First Measurement", "col4", "col5", "Last Measurement")
compare[,c(3,6)]
```



## Are there differences in activity patterns between weekdays and weekends?
---------------------------------------------------------------------------------------------
### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r echo = TRUE}
## add weekend / weekday data
activity.new$week <- as.factor(activity.new$week <- wday(ymd(activity.new[,"date"])))

activity.new$week <- revalue(activity.new$week, c("1"="Weekend", "2"="Weekday", "3"="Weekday", "4"="Weekday", 
                                                  "5"="Weekday", "6"="Weekday", "7"="Weekend"))

## calculate new data set
data.steps.4 <- data.frame(interval=as.numeric(2*length(unique(activity$interval))),
                           steps=as.numeric   (2*length(unique(activity$interval))),
                           week=as.character  (2*length(unique(activity$interval))),
                           stringsAsFactors=FALSE)
count <- 1
for (i in unique(activity.new$interval))
    {        
        data.steps.4[count,"interval"] <- i 
        data.steps.4[count,"steps"]    <- mean(activity.new[activity.new$interval==i & activity.new$week=="Weekday", "steps"])
        data.steps.4[count,"week"]     <- "Weekday"
        count <- count + 1

        data.steps.4[count,"interval"] <- i 
        data.steps.4[count,"steps"]    <- mean(activity.new[activity.new$interval==i & activity.new$week=="Weekend", "steps"])
        data.steps.4[count,"week"]     <- "Weekend"
        count <- count + 1
    }
data.steps.4$week <- as.factor(data.steps.4$week)

str(data.steps.4)
```

---------------------------------------------------------------------------------------------
### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r echo = TRUE}

ggplot(data=data.steps.4, aes(interval, steps)) +
        geom_line(stat="identity") +
        facet_grid(week~.) +
        labs(x = "Interval") +
        labs(y = "Number of steps") +
        labs(title = "Average Number of Steps per 5-minutes time interval") 


```




